#Reproducible Research Project 1
================================
## Nick Rose
## September 2015



First we will setup and load the necessary libraries needed to conduct this 
Analysis
```{r, echo = TRUE}
library(knitr)
library(ggplot2)
opts_chunk$set(echo= TRUE)
```

## Loading and preprocessing the data
The next step is to load in the data, look at it to get a sense of what it
looks like, and preprocess the data so that is useful
```{r data load}
activity <- read.csv("activity.csv")
summary(activity)
activity$date <- as.Date(activity$date)
activity_sub <- na.exclude(activity)
```
In this case, we turned the interval into a factor and converted the dates into date class. We also created a subsetted database with the NA's removed.

## What is mean total number of steps taken per day?
To address this questions ignoring the NAs, we will use the subsetted data to
ge the total number of steps per day. First we will make a new data set compsoed of days and total steps per day
```{r}
steps_per_day <- sapply(split(activity_sub$steps, activity_sub$date), sum)
steps_per_day <- as.data.frame(steps_per_day)
steps_per_day[,2] <- as.factor(rownames(steps_per_day))
rownames(steps_per_day) <- NULL
names(steps_per_day) <- c("Steps", "Day")
summary(steps_per_day)
```

We can see from above that the total steps per day ranges from `r steps_per_day[1,1]` to `r steps_per_day[6,1]` with a median of `r steps_per_day[3,1]` and a mean of   
`r steps_per_day[4,1]`.

We can visualize this data by generating a histogram
```{r}
hist(steps_per_day$Steps, breaks = 20, xlab = "Steps per Day", col = "blue")
```

## What is the average daily activity pattern?
To exmaine the data and compare the total of steps per interval through the day, we will first summarize the steps by interval for all days. 

```{r}
steps_per_interval <- sapply(split(activity_sub$steps, activity_sub$interval), mean)
steps_per_interval <- as.data.frame(steps_per_interval)
steps_per_interval[,2] <- unique(activity_sub$interval)
rownames(steps_per_interval) <- NULL
names(steps_per_interval) <- c("Steps", "Interval")
```

Then we can make a time series plot to visualize this data
```{r}
g <- ggplot(data = steps_per_interval, aes( x =Interval, y = Steps)) + geom_line()
g
```

Then we can look at the data quantitatively 
```{r}
summary(steps_per_interval)
steps_max <- steps_per_interval$Interval[steps_per_interval$Steps == max(steps_per_interval$Steps)]
```
and determine that the time interval during the day that the most steps are taken is `r steps_max`


## Imputing missing values
However, we removed a large amount of NA values. We can also look at the data
making asusmptions about the missing values. We can calculate the number of NA values using the following R code: 
```{r} 
list_NA <- is.na(activity$steps)
num_NAs <- length(list_NA[list_NA == TRUE])
```

The number of NAs in the dataset are `r num_NAs`

To address these missing values, we will use the mean of the steps taken for 
that interval across all days to replace the NA values. The code for this is 
```{r}
sub_NA <- activity[list_NA, ]
for (i in 1:nrow(sub_NA)){
    i_int <- sub_NA[i, 3]
    sub_NA[i, 1] <- steps_per_interval$Steps[steps_per_interval$Interval == i_int]
}
activity_NAs <- rbind(activity_sub, sub_NA)
```
To exmaine how handling the NAs this way impacted our data, we will rerun our analysis of the total steps per day with the new data set.

```{r}
steps_per_day_NA <- sapply(split(activity_NAs$steps, activity_NAs$date), sum)
steps_per_day_NA <- as.data.frame(steps_per_day_NA)
steps_per_day_NA[,2] <- as.factor(rownames(steps_per_day_NA))
rownames(steps_per_day_NA) <- NULL
names(steps_per_day_NA) <- c("Steps", "Day")
summary(steps_per_day_NA)
hist(steps_per_day$Steps, breaks = 20, xlab = "Steps per Day", col = "blue")
```


## Are there differences in activity patterns between weekdays and weekends?
